OptiFit: a fast method for fitting amplicon sequences to existing OTUs

### 2021-03-05 paper/paper.Rmd
Describe benchmarking methods

OptiFit leverages the method employed by OptiClust of iteratively assigning
sequences to OTUs to produce the highest quality OTUs possible, and extends this
method for reference-based clustering.
OptiFit takes as input a list of reference OTUs and their sequences, a list of
query sequences to assign to the reference OTUs, the sequence pairs that are
within the distance threshold (e.g. 0.03), and the metric to assess clustering
quality (default: MCC).
Query sequences are randomly seeded in reference OTUs, then for each sequence
the algorithm calculates the quality metric based on whether the sequence stays
in its current OTU or moves to each of the other OTUs.
This process is repeated until the quality metric stabilizes, changing by no
more than 0.0001 by default, or until a maximum number of iterations is reached
(default: 100).
In closed-reference mode, any query sequences that cannot be assigned are thrown
out (**TODO: exactly what determines whether a seq can't be assigned?**), and
the results only contain OTUs that exist in the original reference.
In open-reference mode, unassigned query sequences are clustered _de novo_ using
OptiClust to generate additional OTUs. The final quality score is reported with
the best OTU assignments.

To evaluate the OptiFit algorithm and compare to existing methods, we used four
published datasets isolated from soil [@johnston_metagenomics_2016], marine
[@henson_artificial_2016], mouse gut [@schloss_stabilization_2012], and human
gut [@baxter_microbiota-based_2016] samples. 
There are two strategies for generating OTUs with OptiFit: 1) fit sequences to
reference OTUs of an independent database, or 2) split the dataset into a
reference and query fraction, cluster the reference sequences _de novo_, then
fit the query sequences to the reference OTUs.
For each dataset repeated with 100 random seeds, we generated OTUs with OptiFit
using both strategies.
To compare to existing software, we also clustered OTUs _de novo_ using
OptiClust and vsearch, and with vsearch in reference-based mode against the
Greengenes database.
All clustering was performed at a distance threshold of 0.03 and OTU quality was
evaluated using the Matthews Correlation Coefficient (MCC) as described
previously [@westcott_opticlust_2017].
We calculated the fraction of query sequences that were fit to existing OTUs in
closed reference mode as an additional measure of quality for this mode.
_De novo_ clustering with OptiClust produced **X%** higher quality OTUs than
vsearch, but performed **Y%** slower than vsearch.

@@

In open reference mode, fitting the datasets to reference OTUs with OptiFit
produced OTUs of similar quality as clustering the datasets _de novo_ with
OptiClust (**diff. in medians**).
This held true for all datasets and reference databases.
OptiFit produced higher quality OTUs than vsearch when open reference clustering
against the Greengenes database, with median MCC scores of **A** and **B%**
(respectively).
However, open reference vsearch and OptiClust respectively ran **X%** and **Y%**
faster than OptiFit in open reference mode.

In closed reference mode, OptiFit produced lower quality OTUs than OptiClust by
**X%** when fitting sequences to Greengenes and SILVA, and **Y%** worse when
fitting to RDP.
Only up to **Z%** of query sequences were fit to reference OTUs in closed
reference mode across any dataset/database combination.
OptiFit was able to fit **X** more query sequences to reference OTUs created
with the Greengenes and SILVA databases than with RDP.
Vsearch was able to fit **W%** more query sequences to the Greengenes reference
than OptiFit in closed reference mode.
In terms of run time, closed reference OptiFit outperformed
OptiClust by **A%** and vsearch by **B%**.

@@

The OptiClust and OptiFit algorithms provided by mothur produced higher quality
OTUs than vsearch in open reference, closed reference, or _de novo_ modes.
However, vsearch was able to fit more sequence into OTUs than OptiFit in closed
reference mode.
While both mothur and vsearch use a distance or similarity threshold for
determining how to assign sequences to OTUs, vsearch is more permissive than
mothur.
Mothur's OptiFit and OptiClust algorithms use all of the sequences to define an
OTU, requiring that all pairs of sequences (including reference and query
sequences) in an OTU are within the distance threshold without penalizing the
MCC.
In contrast, vsearch only requires each query sequence to be similar to the
single sequence that seeded the OTU.
In this way, vsearch sacrifices OTU quality in order to allow more sequences to
fit to OTUs.
Users who require closed reference clustering to the Greengenes database may
prefer to use vsearch if they wish to maximize the fraction of sequences that
can be fit at the cost of OTU quality.
However, mothur's OptiClust or OptiFit are recommended for _de novo_ or open
reference clustering to produce OTUs of the highest possible quality.

@@

**TODO: big picture concluding paragraph.**
We have developed a new reference-based clustering algorithm that allows users
to produce high quality OTUs when stability is a requirement for the task at hand.

@@

To generate reference OTUs from independent databases, we downloaded sequences
from the Greengenes database (v13_8_99) [@desantis_greengenes_2006], SILVA
non-redundant database (v132) [@quast_silva_2013], and the Ribosomal Database
Project (v16) [@cole_ribosomal_2014].
These sequences were processed using the same steps outlined above followed by
clustering sequences into _de novo_ OTUs with OptiClust.
Processed reads from each of the four datasets were clustered with OptiFit to
the reference OTUs generated from each of the three databases.
When reference clustering with vsearch, processed datasets were fit directly to
the unprocessed Greengenes reference alignment, since this method is how vsearch
is typically used by the QIIME2 software reference-based clustering
[@bolyen_reproducible_2019; @noauthor_clustering_nodate].

@@

**TODO: describe split dataset creation**

@@

@@

We implemented the analysis workflow in Snakemake [@koster_snakemake_2012] and
wrote scripts in R [@r_core_team_r_2020], Python [@van_rossum_python_2009], and
GNU bash [@noauthor_bash_nodate].
Software used includes mothur v1.45.0 [@schloss_introducing_2009], vsearch
v2.13.3 [@rognes_vsearch_2016], numpy [@harris_array_2020], the Tidyverse
metapackage [@wickham_welcome_2019], R Markdown [@xie_r_2018], the SRA toolkit
[@noauthor_sra-tools_nodate], and the conda environment manager
[@noauthor_anaconda_2016].
The complete workflow, manuscript, and conda environment are available at
**TODO: UPDATED REPO LINK**.

>>>>

OptiFit leverages the method employed by OptiClust of iteratively assigning
sequences to OTUs to produce the highest quality OTUs possible, and extends this
method for reference-based clustering.
OptiFit takes as input a list of reference OTUs and their sequences, a list of
query sequences to assign to the reference OTUs, the sequence pairs that are
within the distance threshold (e.g. 0.03), and the metric to assess clustering
quality (default: MCC).
Query sequences are randomly seeded in reference OTUs, then for each sequence
the algorithm calculates the quality metric based on whether the sequence stays
in its current OTU or moves to each of the other OTUs.
If two or more OTU assignments are of equal quality, a random number generator
is used to break the tie.
This process is repeated until the quality metric stabilizes, changing by no
more than 0.0001 by default, or until a maximum number of iterations is reached
(default: 100).
In closed-reference mode, any query sequences that cannot be assigned are thrown
out (**TODO: exactly what determines whether a seq can't be assigned?**), and
the results only contain OTUs that exist in the original reference.
In open-reference mode, unassigned query sequences are clustered _de novo_ using
OptiClust to generate additional OTUs. The final quality score is reported with
the best OTU assignments.

To evaluate the OptiFit algorithm and compare to existing methods, we used four
published datasets isolated from soil [@johnston_metagenomics_2016], marine
[@henson_artificial_2016], mouse gut [@schloss_stabilization_2012], and human
gut [@baxter_microbiota-based_2016] samples. 
There are two strategies for generating OTUs with OptiFit: 1) fit sequences to
reference OTUs of an independent database, or 2) split the dataset into a
reference and query fraction, cluster the reference sequences _de novo_, then
fit the query sequences to the reference OTUs.
For each dataset repeated with 100 random seeds, we generated OTUs with OptiFit
using both strategies.
To compare to existing software, we also clustered OTUs _de novo_ using
OptiClust and VSEARCH, and with VSEARCH in reference-based mode against the
Greengenes database.
All clustering was performed at a distance threshold of 0.03 and OTU quality was
evaluated using the MCC as described previously [@westcott_opticlust_2017].
We calculated the fraction of query sequences that were fit to existing OTUs in
closed reference mode as an additional measure of quality for this mode.
_De novo_ clustering with OptiClust produced **X%** higher quality OTUs than
VSEARCH, but performed **Y%** slower than VSEARCH.

@@

In open reference mode, fitting the datasets to reference OTUs with OptiFit
produced OTUs of similar quality as clustering the datasets _de novo_ with
OptiClust (**diff. in medians**).
This held true for all datasets and reference databases.
OptiFit produced higher quality OTUs than VSEARCH when open reference clustering
against the Greengenes database, with median MCC scores of **A** and **B%**
(respectively).
However, open reference VSEARCH and OptiClust respectively ran **X%** and **Y%**
faster than OptiFit in open reference mode.

In closed reference mode, OptiFit produced lower quality OTUs than OptiClust by
**X%** when fitting sequences to Greengenes and SILVA, and **Y%** worse when
fitting to RDP.
Only up to **Z%** of query sequences were fit to reference OTUs in closed
reference mode across any dataset/database combination.
OptiFit was able to fit **X** more query sequences to reference OTUs created
with the Greengenes and SILVA databases than with RDP.
VSEARCH was able to fit **W%** more query sequences to the Greengenes reference
than OptiFit in closed reference mode.
In terms of run time, closed reference OptiFit outperformed
OptiClust by **A%** and VSEARCH by **B%**.

@@

The OptiClust and OptiFit algorithms provided by mothur produced higher quality
OTUs than VSEARCH in open reference, closed reference, or _de novo_ modes.
However, VSEARCH was able to fit more sequence into OTUs than OptiFit in closed
reference mode.
While both mothur and VSEARCH use a distance or similarity threshold for
determining how to assign sequences to OTUs, VSEARCH is more permissive than
mothur.
Mothur's OptiFit and OptiClust algorithms use all of the sequences to define an
OTU, requiring that all pairs of sequences (including reference and query
sequences) in an OTU are within the distance threshold without penalizing the
MCC.
In contrast, VSEARCH only requires each query sequence to be similar to the
single sequence that seeded the OTU.
In this way, VSEARCH sacrifices OTU quality in order to allow more sequences to
fit to OTUs.
Users who require closed reference clustering to the Greengenes database may
prefer to use VSEARCH if they wish to maximize the fraction of sequences that
can be fit at the cost of OTU quality.
However, mothur's OptiClust or OptiFit are recommended for _de novo_ or open
reference clustering to produce OTUs of the highest possible quality.

@@

**TODO: big picture concluding paragraph.**
We have developed a new clustering algorithm that allows users to produce high
quality OTUs using already existing OTUs as a reference.

@@

To generate reference OTUs from independent databases, we downloaded sequences
from the Greengenes database (v13_8_99) [@desantis_greengenes_2006], SILVA
non-redundant database (v132) [@quast_silva_2013], and the Ribosomal Database
Project (v16) [@cole_ribosomal_2014].
These sequences were processed using the same steps outlined above followed by
clustering sequences into _de novo_ OTUs with OptiClust.
Processed reads from each of the four datasets were clustered with OptiFit to
the reference OTUs generated from each of the three databases.
When reference clustering with VSEARCH, processed datasets were fit directly to
the unprocessed Greengenes reference alignment, since this method is how VSEARCH
is typically used by the QIIME2 software reference-based clustering
[@bolyen_reproducible_2019; @noauthor_clustering_nodate].

@@

For each dataset, a fraction of the sequences was selected to be clustered _de
novo_ into reference OTUs with OptiClust.
We used three methods for selecting the fraction of sequences to be used
as the reference; a simple random sample, weighting sequences by relative
abundance, and weighting by similarity to other sequences in the dataset.
Dataset splitting was repeated with reference fractions ranging from 10% to 80%
of the dataset and for 100 random seeds.
For each dataset split, the remaining sequences were assigned to the reference
OTUs with OptiFit.

@@

Since OptiClust and OptiFit employ a random number generator to break ties
when OTU assignments are of equal quality, they produce slightly different OTU
assignments when repeated with different random seeds.
To capture any variation in OTU quality or execution time, clustering was
repeated with 100 random seeds for each combination of parameters and
input datasets.
We used the benchmark feature provided by Snakemake to measure the run time of
every clustering job.
We calculated the MCC on each set of OTUs to quantify the quality of clustering,
as described by Westcott _et al._ [@westcott_opticlust_2017].

@@

We implemented the analysis workflow in Snakemake [@koster_snakemake_2012] and
wrote scripts in R [@r_core_team_r_2020], Python [@van_rossum_python_2009], and
GNU bash [@noauthor_bash_nodate].
Software used includes mothur v1.45.0 [@schloss_introducing_2009], VSEARCH
v2.13.3 [@rognes_vsearch_2016], numpy [@harris_array_2020], the Tidyverse
metapackage [@wickham_welcome_2019], R Markdown [@xie_r_2018], the SRA toolkit
[@noauthor_sra-tools_nodate], and the conda environment manager
[@noauthor_anaconda_2016].
The complete workflow, manuscript, and conda environment are available at
**TODO: UPDATED REPO LINK**.
